---
title: "Hands-on_Ex09"
author: "Ding Hong Yan"
date-modified: "last-modified" 
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## **9.1 Overview**

Ternary plots are a way of displaying the distribution and variability of three-part compositional data. (For example, the proportion of aged, economy active and young population or sand, silt, and clay in soil.) It’s display is a triangle with sides scaled from 0 to 1. Each side represents one of the three components. A point is plotted so that a line drawn perpendicular from the point to each leg of the triangle intersect at the component values of the point.

In this hands-on, you will learn how to build ternary plot programmatically using R for visualising and analysing population structure of Singapore.

The hands-on exercise consists of four steps:

-   Install and launch **tidyverse** and **ggtern** packages.

-   Derive three new measures using *mutate()* function of **dplyr** package.

-   Build a static ternary plot using *ggtern()* function of **ggtern** package.

-   Build an interactive ternary plot using *plot-ly()* function of **Plotly R** package.

## **9.2 Installing and launching R packages**

For this exercise, two main R packages will be used in this hands-on exercise, they are:

-   [**ggtern**](http://www.ggtern.com/), a ggplot extension specially designed to plot ternary diagrams. The package will be used to plot static ternary plots.

-   [**Plotly R**](https://plot.ly/r/), an R package for creating interactive web-based graphs via plotly’s JavaScript graphing library, plotly.js . The **plotly R** libary contains the *ggplotly* function, which will convert **ggplot2** figures into a Plotly object.

We will also need to ensure that selected **tidyverse** family packages namely: **readr**, **dplyr** and **tidyr** are also installed and loaded.

In this exercise, version 3.2.1 of **ggplot2** will be installed instead of the latest version of **ggplot2**. This is because the current version of **ggtern** package is not compatible to the latest version of **ggplot2**.

The code chunks below will accomplish the task.

```{r}
pacman::p_load(plotly, ggtern, tidyverse)
```

## **9.3 Data Preparation**

### **9.3.1 The data**

For the purpose of this hands-on exercise, the [Singapore Residents by Planning AreaSubzone, Age Group, Sex and Type of Dwelling, June 2000-2018](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) data will be used. The data set has been downloaded and included in the data sub-folder of the hands-on exercise folder. It is called *respopagsex2000to2018_tidy.csv* and is in csv file format.

### **9.3.2 Importing Data**

To important *respopagsex2000to2018_tidy.csv* into R, ***read_csv()*** function of **readr** package will be used.

```{r}
pop_data <- read_csv("data/respopagsex2000to2018_tidy.csv") 
```

### **9.3.3 Preparing the Data**

Next, use the ***mutate()*** function of **dplyr** package to derive three new measures, namely: young, active, and old.

```{r}
agpop_mutated <- pop_data %>%
  mutate(`Year` = as.character(Year))%>%
  spread(AG, Population) %>%
  mutate(YOUNG = rowSums(.[4:8]))%>%
  mutate(ACTIVE = rowSums(.[9:16]))  %>%
  mutate(OLD = rowSums(.[17:21])) %>%
  mutate(TOTAL = rowSums(.[22:24])) %>%
  filter(Year == 2018)%>%
  filter(TOTAL > 0)
```

::: callout-note
## spread(AG, Population)

The `spread()` function expands the values in the **AG** column into multiple new columns, using **Population** as the corresponding values.

**Example: Before Transformation (`pop_data` original format)**

| Year | AG    | Population |
|------|-------|------------|
| 2018 | 0-4   | 50000      |
| 2018 | 5-9   | 52000      |
| 2018 | 10-14 | 53000      |

**After Transformation (`spread()` or `pivot_wider()`)**

| Year | 0-4   | 5-9   | 10-14 |
|------|-------|-------|-------|
| 2018 | 50000 | 52000 | 53000 |
:::

::: callout-note
## mutate(YOUNG = rowSums(.\[4:8\]))

-   `.[4:8]` refers to columns **4 to 8**, which should correspond to age groups **0-4 to 20-24 years**.

-   `rowSums()` computes the sum across these columns, representing the total population of the young age group.
:::

## **9.4 Plotting Ternary Diagram with R**

### **9.4.1 4.1 Plotting a static ternary diagram**

Use ***ggtern()*** function of **ggtern** package to create a simple ternary plot.

```{r}
ggtern(data=agpop_mutated,aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point()
```

```{r}
ggtern(data=agpop_mutated, aes(x=YOUNG,y=ACTIVE, z=OLD)) +
  geom_point() +
  labs(title="Population structure, 2015") +
  theme_rgbw()
```

### **9.4.2 Plotting an interative ternary diagram**

The code below create an interactive ternary plot using ***plot_ly()*** function of **Plotly R**.

```{r}
#| message: false
#| warning: false
label <- function(txt) {
  list(
    text = txt, 
    x = 0.1, y = 1,
    ax = 0, ay = 0,
    xref = "paper", yref = "paper", 
    align = "center",
    font = list(family = "serif", size = 15, color = "white"),
    bgcolor = "#b3b3b3", bordercolor = "black", borderwidth = 2
  )
}

# reusable function for axis formatting
axis <- function(txt) {
  list(
    title = txt, tickformat = ".0%", tickfont = list(size = 10)
  )
}

ternaryAxes <- list(
  aaxis = axis("Young"), 
  baxis = axis("Active"), 
  caxis = axis("Old")
)

# Initiating a plotly visualization 
plot_ly(
  agpop_mutated, 
  a = ~YOUNG, 
  b = ~ACTIVE, 
  c = ~OLD, 
  color = I("black"), 
  type = "scatterternary"
) %>%
  layout(
    annotations = label("Ternary Markers"), 
    ternary = ternaryAxes
  )
```

::: callout-important
## Ternary Plot Explanation

A **Ternary Plot** is a visualization tool for displaying **three interrelated variables** whose total sum is **fixed at 1 (or 100%)**. It is widely used in fields such as **geology, chemistry, economics, and ecology** to analyze proportional compositions.

**1. Structure of a Ternary Plot**

A ternary plot is an **equilateral triangle**, where each **vertex represents a variable (A, B, C)**. Each **point inside the triangle** represents a combination of the three variables. For example:

-   **Vertex A (100%, 0%, 0%)**

-   **Vertex B (0%, 100%, 0%)**

-   **Vertex C (0%, 0%, 100%)**

This type of plot is suitable for data where the **total proportion is always 100% (or 1.0)**.

**2. How to Analyze a Ternary Plot**

1.  **Trend Analysis**

    -   If data points cluster near a particular **vertex**, it indicates that this variable has a **high proportion**.

    -   If data points are evenly distributed, the three variables are relatively balanced.

2.  **Clustering and Patterns**

    -   Look for **clusters** of data points, which can indicate different groups.

    -   Useful for **market segmentation, rock classification,** and other analytical applications.

3.  **Impact of Mixtures**

    -   Analyze how different component ratios affect outcomes, such as **the effect of cement composition on strength**.

## **Summary**

-   **Ternary plots** are used for **three variables that sum to 1 (or 100%)**.

-   **Each point inside the triangle** represents a different combination of the three variables.

-   **Applications** include **market analysis, chemistry, geology, and ecology**.

-   **Analysis methods** include identifying trends, clusters, and the influence of different proportions.
:::
