---
title: "Take-home_Ex01"
author: "Ding Hong Yan"
date-modified: "last-modified" 
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Overview

### **Setting the scene**

An international media company that publishes weekly content on digital platforms is planning to release articles on the following themes:

-   Ship Performance in the Gulf of Guinea

### **The Task**

In this exercise, Exploratory Data Analysis (EDA) methods and ggplot functions are used to explore:

-   the relationship between ship performance with Ship_Type, Engine_Type, Engine_Power_kW, and Weather_Condition.

-   the operating costs and revenues of different route types to identify the most profitable route.

## Getting started

We load the following R packages using the `pacman::p_load()` function:

-   **tidyverse**: Core collection of R packages designed for data science

-   **ggrepel**: to provides geoms for **ggplot2** to repel overlapping text labels

-   **ggthemes**: to use additional themes for **ggplot2**

-   **patchwork**: to prepare composite figure created using **ggplot2**

-   **ggridges**: to plot ridgeline plots

-   **ggdist**: for visualizations of distributions and uncertainty

-   **scales**: provides the internal scaling infrastructure used by **ggplot2**

```{r}
pacman::p_load(tidyverse, haven,
               ggrepel, ggthemes,
               ggridges, ggdist,
               patchwork, scales)
```

### Import data

In this project, Ship Performance in the Gulf of Guinea will be used. **The data was downloaded from [kaggle website](https://www.kaggle.com/datasets/jeleeladekunlefijabi/ship-performance-clustering-dataset?resource=download).**

**The code chunk below imports the data into R and save it into a tibble data frame object called *ship_pf*.**

```{r}
ship_pf <- read_csv("data/Ship_Performance_Dataset.csv")
```
